A Data-Driven Framework for Neural Field Modeling

### 2011-01-04 ltx/first_order_framework/response_to_reviews.tex

        We first address the reviewer's main concerns, followed by the reviewer's additional comments.

    \subsection{Main Concerns}
    It would be useful to mention the above (now below) explicitly in the introduction and try to address them in later sections of the paper.

    \begin{enumerate}
        \item Do not use real data for the validation of their approach

	\emph{The authors appreciate that the only way to truly validate the proposed framework is with real data. However, before validating the method on real data we believe that it is essential to first validate the estimation method using accepted models and synthetic data, where the true and estimated parameters can be compared. In an effort to make the method employed in the paper more explicit, we have added ``To illustrate the estimation framework, data is generated using the neural field equations incorporating modeled sensors enabling a comparison between the estimated and true parameters'' to the introduction. Another paper demonstrating the application of the method with data recorded from a Utah array implanted in an epilepsy surgical patient is currently in preparation.  WE SHOULD ADD SOMETHING IN THE CONCLUSION TO SAY WE TOTALLY NEED TO VALIDATE THIS ON REAL DATA AND IT'S NOTED AS FUTURE WORK}

@@

        \item The use of Euler discretisation and how this affects the system's behavior.

		\emph{This point was made by a reviewer from our PLoS attempt. The reviewer stated: ``1. The authors might want to reference prior work in using Kalman
		Filtering for the estimation of neural mass models from data [1-3],
		[13]. In fact from those papers stressed the importance of carrying out
		a high order discretization of the differential equations, which should
		be in fact treated as stochastic. In equation (S1.1) of the extended
		derivation, it is shown that an Euler approximation is taken which may
		even change the dynamical properties of the system. The authors should
		discuss this issue.'' ALL THESE PAPERS USE a method called local linearisation approach for discretisation.}

        \item The derivation of Equations (30) and (31) [spacing of basis functions] and make them more transparent for the general audience.

\emph{This appendix provides a derivation for equations ?? and ??. The derivation is adapted from (Sanner et al., 1992, and Scerri et al., 2009). Consider the cortical sheet defined on the surface $\Omega$ with the sensors placed on a regular grid at positions $\mathbf{r}_n$ (where $n=0,...,n_y-1$ indexes the sensors) that are separated by the distance $\Delta_y$. The observations can be considered samples from a continuous field that is filtered (convolved) with the observation kernel described by
\begin{equation}
	y(\mathbf{r}_n)=\sum_{n=0}^{n_y-1}\delta(\mathbf r - \mathbf{r}_n ) v_m(\mathbf{r}),
\end{equation} 
where $v_m(\mathbf{r}) = v_t(\mathbf{r})\ast m(\mathbf{r})$. The Fourier transform of $y(\mathbf{r}_n)$ is
\begin{align}
 Y(\boldsymbol{\nu}) &= \frac{1}{\Delta_y^2}\sum_{n=0}^{n_y-1} \delta\left(\boldsymbol{\nu}-\frac{n}{\Delta_y}\right) \ast V_m(\boldsymbol \nu) \\
	&= \frac{1}{\Delta_y^2}V_m(\boldsymbol\nu)+\frac{1}{\Delta_y^2}\sum_{n=1}^{n_y-1} V_m\left(\boldsymbol{\nu}-\frac{n}{\Delta_y}\right).
\end{align} 
It should be clear that $Y(\boldsymbol{\nu})$ contains $V_m(\boldsymbol\nu)$ plus $n_y-2$ copies (aliases) of $V_m(\boldsymbol\nu)$ repeated on the regular lattice centered at $n\Delta_y^{-1}$. Due to the lowpass action of the observation kernel, the filtered dynamic field is band-limited with a cutoff frequency $\boldsymbol\nu_{cy}$. In order for $V_m(\boldsymbol\nu)$ to be disjointed from its aliases, the amount of shift $\Delta_y^{-1}$ must be at least $2\boldsymbol\nu_{cy}$, since each alias has a spectral range of
\begin{equation}
	\boldsymbol\nu = \left[\frac{n}{\Delta_y}-\boldsymbol{\nu}_{cy},\frac{n}{\Delta_y}+\boldsymbol{\nu}_{cy}\right].
\end{equation}
Therefore, to avoid aliasing the distance between observations must satisfy
\begin{equation}\label{eq:SamplingCriteria}
	\frac{1}{\Delta_y} \ge 2\nu_{cy}.
\end{equation}
Since the observation kernel is not an ideal filter an oversampling parameter should be included to further separate the aliases. Rearranging equation~\ref{eq:SamplingCriteria} and adding the oversampling parameter, $\rho$, gives
\begin{equation}\label{ap:Shannon}
	\Delta_y\le\frac{1}{2\rho\nu_{cy}},
\end{equation}
where $\rho \in \mathbb{R} \ge 1$. }

        \item The existing literature on inversion schemes for similar models of brain activity.
        \item In section 3, Table 6 is mentioned however there is no such table in the ms.
    \end{enumerate}

@@

    \section{Reviewer 2}

    We thank the author for their encouraging review. Each point raised is dealt with in turn.

\begin{enumerate}
    \item The handling of the (to borrow a term from numerical weather forecasting) `sub-grid scale physics' is inadequate for nervous systems (see, e.g., equation 8 [The nonlinear IntegroDifference Equation]). I do very much appreciate the spectral estimation of spatial and temporal scales. But the model errors are lumped into an iid model error term that is not realistic. For instance, the flow of potassium in the extracellular spaces has a known dynamic that will affect network excitability, generating important sub-grid dynamics that is anything but iid. See, for instance, the example in Ullah et al PLoS Comp Biol 2010 where without such dynamics, state tracking and parameter estimation for a neural system fails. A beautiful overview of sub-grid effects in data assimilation is found in Kalnay's 2003 book. I would very much value the authors comments on these issues. I guess were I to respond to this question, that I might consider extending the model error term to an autoregressive model. But I will let the authors answer this question.\\

    \emph{talk about model order}

	\emph{Indeed, there will always be a mismatch between a mathematical model and the natural phenomena it is attempting to describe. The model proposed in our paper is modest in terms of the temporal dynamics, where the state transition is assumed to be first-order. However, the development of the framework for this modest brain model is a significant development that can be expanded to include longer range time dependencies. A major benefit in using a mean field model is the freedom to lump detailed parameters, such as specific ion concentrations, into more general parameters such as the connectivity kernel coefficients that describe the net system gain. Please note, we have assumed that factors influencing excitability are stationary over the estimation period. This is a reasonable assumption is applying the framework to an evoked potential experiment. An extension to our framework, that is currently in development, is to allow for time-varying parameter estimates. This will allow the application of the framework to longer-term data recordings. The problem then can be posed in terms of convergence time and estimation accuracy vs. nonstationarity of the metabolic processes and other unmodeled brain dynamics.}

@@

\item Although we all know the 1970s version of Wilson and Cowan's field equations, there was an intriguing stochasticization of these equations by Benayoun et al PloS Comp Biol 2010. I and others will value the authors' thoughts of how a more palatable stochasticization of these field equations might influence the future development of the data-assimilation framework presented in this paper.

\item This is NeuroImage, not the Physical Review. Although most of this framework involves mathematics that is reasonable to follow for typical readers of this journal, I would suggest that the authors explain and expand a bit on the introduction of Green's functions where introduced in equation 6.\\

\emph{The synaptic response kernel can be viewed as the Green function of a linear differential equation defined by the differential operator $\textrm{D}=d/dt + \zeta$ i.e., satisfying
\begin{equation}
	\label{GreensFuncDef} \textrm{D}h\left( t \right) = \delta \left( t \right), 
\end{equation} 
where $\delta(t)$ is the Dirac-delta function. Applying the differential operator $\textrm{D}$ on equation (5) we have
\begin{align}
 \textrm{D}v\left(\mathbf r,t\right)&= \textrm{D}\left(h\ast g\right)\left(\mathbf r,t\right)\\
&=\left(\textrm{D}h\ast g\right)\left(\mathbf r,t\right)\\
&=\left(\delta \left(t\right)\ast g\right)\left(\mathbf r,t\right)\\
&=g\left(\mathbf r,t\right)
\end{align}
where $\ast$ denotes convolution operator. This gives the standard form of the model
\begin{equation}
	\label{FinalFormContinuous} 
	\frac{dv\left( \mathbf{r},t \right)}{dt} + \zeta v\left( \mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( {v\left( \mathbf{r}',t \right)} \right)\, d\mathbf{r}'}. 
\end{equation}}

\item What is the relation of the substitution of the temporal and spatial basis functions as employed in this paper to Galerkin's projection? How do the two techniques differ? If this is close, the non-local case for a Galerkin projection (typically used in diffusive processes) has not, to my knowledge, been done. Certainly not in a way that is obvious for neural field theory. Perhaps this is my ignorance. This seems important to clarify for the rest of us.
\item In the isotropic and translationally invariant case examined here, would a Fourier basis for temporal and spatial structure have been a formally correct alternative?\\

\emph{Gaussians has semi-compact support but Fourier basis functions are global functions. Gaussian can be easily extended to higher dimension (don't know about the Fourier basis functions), any orthonormal basis functions like Fourier basis functions simplify our equations, $\Gamma$ becomes identity matrix, but using Gaussian basis functions equations are in their generic form.}

\item In equation 29, what would a reasonable $v_c$ be for a real neuronal system (mammalian) given the number of neurons and scale of electrodes? I am concerned that it might be much higher than any of us wish to consider. One would perhaps be saved by the filter characteristics of brain (e.g. Phys. Rev. E 73, 051911 (2006)).
\item The two covariance inflation parameters in equation 41 make me cringe. How are we best to adjust these fudge factors?
\emph{The positive constant $\alpha$ determines the spread of the sigma points around $\mathbf{\bar x}$ which can be made arbitrary small (e.g. $0<\alpha\le 10^{-4} $) to minimise higher order effect (cite Julier (The Scaled Unscented Transformation) and  Haykin (Kalman filtering and neural network) , and  $\kappa$ is an additional scaling parameter which is usually set to $3-n_x$. cite Julier A New Approach for Filtering Nonlinear Systems}

\item I understand that the Mexican Hat bases are in this paper often a composite of $\theta_0$, $\theta_1$, and $\theta_2$, representing short range excitation, short range inhibition, and long range excitation. But many will be confused by this subtlety, and at times in the text the Mexican Hat basis is described as if it were a single function. I would show how the theta's sum to create the general spatial basis shape of the hat in figure 1, and go through the paper in order to clarify this issue.\\
\emph{I think we need to show each kernel basis function in Figure 1}
 \begin{figure*}[!ht]
 \begin{center}
 \includegraphics{./Graph/pdf/fig1_revised.pdf} 
 \end{center}
 \caption{}
 \label{fig:Figure1}
 \end{figure*}

### 2011-01-04 ltx/first_order_framework/response_to_reviews.tex

        \item The derivation of Equations (30) and (31) [spacing of basis functions] and make them more transparent for the general audience.

\emph{To address this point, we have added an extra appendix with a derivation of the sampling theorem with the appropriate cross-referencing to the main text. An effort was made to keep this derivation minimal and as clear as possible. We have also added the references to relevant part of the text (that should have been there before, as the reviewer has correctly pointed out). We felt that it was worth adding the extra appendix to provide a clear derivation that is contextually and notationally relevant to this paper, as the derivations cited papers may not be clear to the Neuroimage general audience.}

        \item The existing literature on inversion schemes for similar models of brain activity.

	\emph{Valdes uses a Lopes Da Silva / Freeman neural mass model. }

@@

\item In the isotropic and translationally invariant case examined here, would a Fourier basis for temporal and spatial structure have been a formally correct alternative?\\

\emph{Gaussians have semi-compact support but Fourier basis functions are global functions. Gaussian can be easily extended to higher dimension (don't know about the Fourier basis functions), any orthonormal basis functions like Fourier basis functions simplify our equations, $\Gamma$ becomes identity matrix, but using Gaussian basis functions equations are in their generic form.}

\item In equation 29, what would a reasonable $v_c$ be for a real neuronal system (mammalian) given the number of neurons and scale of electrodes? I am concerned that it might be much higher than any of us wish to consider. One would perhaps be saved by the filter characteristics of brain (e.g. Phys. Rev. E 73, 051911 (2006)).

\emph{The spatial frequency content is seems to be actually lower than one might first think. We have added `From the work of (Freeman et al, 2000), the expected bandwidth of the human neocortex is approximately 0.1-0.4 cycles/mm which requires a sensor spacing of approximately 1.25 mm to avoid distortions due to aliasing' to the Frequency Analysis section of the paper to clear this up. We have also added to the discussion. Note, we expect the spatial bandwidth of the cortex to be governed by the connectivity kernel, since it is equivalent to a spatial low-pass filter acting on the firing rates.  }

@@

\item The two covariance inflation parameters in equation 41 make me cringe. How are we best to adjust these fudge factors?\\

\emph{Values for the two parameters in equation 41 along with the relevant references ,that describe the rationale for parameter choice, have been added to the manuscript.}

\item I understand that the Mexican Hat bases are in this paper often a composite of $\theta_0$, $\theta_1$, and $\theta_2$, representing short range excitation, short range inhibition, and long range excitation. But many will be confused by this subtlety, and at times in the text the Mexican Hat basis is described as if it were a single function. I would show how the theta's sum to create the general spatial basis shape of the hat in figure 1, and go through the paper in order to clarify this issue.\\

\emph{This point is clarified by adding the connectivity kernel components superimposed by the resulting Mexican Hat basis in Figure 1. of the manuscript.}

@@

 \begin{figure*}[!ht]
 \begin{center}
 \includegraphics{./Graph/pdf/fig1.pdf} 
 \end{center}
 \caption{}
 \label{fig:Figure1}
 \end{figure*}

\item A few comments on how to adapt the smoother for real time implementation would be helpful. I presume that one would perform the smoothing during a training phase. But one would also wish to adapt in real time as observation or control scenarios progressed. A few words of perspective would be invaluable to help others translate the author's wisdom on this method to other applications.

\emph{A few comments have been added to the Modeling Assumptions section of the discussion.}

\item Figure 4. Should the axes be indicated as spatial frequency rather than just Hz?

### 2011-01-04 ltx/first_order_framework/LaTeX_Manuscript.tex
fixed table referencing reviewrone comment 5
small edits

In practice, it is difficult to estimate the bandwidth of the cortex using traditional electrophysiological measurements, possibly preventing placement of sensors in accordance to equation~\ref{eq:MinimumSensorDistance}. However, we envisage it may be possible to estimate the spatial bandwidth using other modalities with higher spatial resolution such as fMRI, NIRS or other optical imaging techniques (\cite{Issa2000}). Spectral aliasing can be avoided by a proper choice of the spatial sampling distance given the sensors' spectral characteristics. The spatial extent of the sensors results in a spectral low-pass action, thus providing spatial anti-aliasing filtering. Such sensors limit the spatial bandwidth of the observation to a value denoted by $\boldsymbol{\nu}_{cy}$. Aliasing can then be avoided by positioning the sensors according to equation~\ref{eq:MinimumSensorDistance}, replacing $\boldsymbol{\nu}_c$ with $\boldsymbol{\nu}_{cy}$.

@@

The additive form of the unscented Rauch-Tung-Striebel smoother (URTSS) \citep{Sarkka2010} was used for the state estimation. The URTSS incorporates an unscented Kalman filter (UKF) \citep{Julier1997, Merwe2003} in a forward step to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward step to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. The first and second order moments of the predicted state are captured by propagating the so-called sigma points through the state equation. The sigma points, $\mathcal X_i$, are calculated using the unscented transform as follows:
\begin{align}\label{eq:sigmapoints1}
	\mathcal X_{0}&=\mathbf{\bar x} \\
	\mathcal X_{i}&= \mathbf{\bar x}+\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_i, \quad i=1, \dots, n_x \\
	\mathcal X_{i}&=\mathbf{\bar x}-\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_{i- n_x}, \quad i= n_x+1, \dots, 2n_x 
\end{align}
where $\mathbf{\bar x}$ represents either $\hat{\mathbf x}_t^{f}$ or $\hat{\mathbf x}_t^{b}$, $\mathbf{P}_x$ is the corresponding covariance matrix for filtering or smoothing, $\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_i$ is the $i^{th}$ column of the scaled matrix square root of $\mathbf P_x$, and $n_x$ is the dimension of the state space. The total number of sigma points is $2n_x+1$. The scaling parameter, $\lambda$, is defined as 
\begin{equation}\label{eq:sigmapoints3}
	\lambda=\alpha^2( n_x+\kappa) - n_x, 
\end{equation}
\parham{(reviewer 2, comment 7:) The positive constant $\alpha$ determines the spread of the sigma points around $\mathbf{\bar x}$ which is set arbitrary small (e.g., $10^{-3}$) to minimise higher order effect \citep{Haykin2001} , and  $\kappa$ is an additional scaling parameter which is usually set to $3-n_x$ \citep{Julier2002a}.}

The sigma vectors are propagated through the system equations and weighted to form the predicted mean and covariance. The weights are calculated by 
\begin{align}
	\mathbf W_0^{(m)}&=\frac{\lambda}{ n_x+\lambda} \\
	\mathbf W_0^{(c)}&=\frac{\lambda}{ n_x+\lambda}+(1-\alpha^2+\beta) \\
	\mathbf W_i^{(m)}&=\mathbf W_i^{(c)}=\frac{1}{2( n_x+\lambda)} \quad i=1, \dots, 2n_x, 
\end{align}
where the superscripts $m$ and $c$ denote mean and covariance and $\beta$ incorporates prior knowledge of the distribution of the states, $\mathbf{x}$. Since the observation equation is linear (equation~\ref{ObservationEquation}), the standard Kalman Filter update equations are used to correct the predicted states. The state estimates from forward filtering are used to form a new set of sigma points for the smoother, as described above. The cross-covariance matrix of the states, $\mathbf M$, is also required for computing the smoother gain, as described in the summary of the steps in the URTSS algorithm is given in Table~\ref{tab:UKFAlgorithm}.

@@

>>>>

In practice, it is difficult to estimate the bandwidth of the cortex using traditional electrophysiological measurements, possibly preventing placement of sensors in accordance to equation~\ref{eq:MinimumSensorDistance}. However, we envisage it may be possible to estimate the spatial bandwidth using other modalities with higher spatial resolution such as fMRI, NIRS or other optical imaging techniques (\cite{Issa2000}). \dean{(Reviewer 2, comment 6:) From the work of~\citet{Freeman2000}, the expected bandwidth of the human neocortex is approximately 0.1-0.4 cycles/mm.}

Spectral aliasing can be avoided by a proper choice of the spatial sampling distance given the sensors' spectral characteristics. The spatial extent of the sensors results in a spectral low-pass action, thus providing spatial anti-aliasing filtering. Such sensors limit the spatial bandwidth of the observation to a value denoted by $\boldsymbol{\nu}_{cy}$. Aliasing can then be avoided by positioning the sensors according to equation~\ref{eq:MinimumSensorDistance}, replacing $\boldsymbol{\nu}_c$ with $\boldsymbol{\nu}_{cy}$.

@@

The additive form of the unscented Rauch-Tung-Striebel smoother (URTSS) \citep{Sarkka2010} was used for the state estimation. The URTSS incorporates an unscented Kalman filter (UKF) \citep{Julier1997, Merwe2003} in a forward step to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward step to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. The first and second order moments of the predicted state are captured by propagating the so-called sigma points through the state equation. The sigma points, $\mathcal X_i$, are calculated using the unscented transform as follows:
\begin{align}\label{eq:sigmapoints1}
	\mathcal X_{0}&=\mathbf{\bar x} \\
	\mathcal X_{i}&= \mathbf{\bar x}+\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_i, \quad i=1, \dots, n_x \\
	\mathcal X_{i}&=\mathbf{\bar x}-\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_{i- n_x}, \quad i= n_x+1, \dots, 2n_x 
\end{align}
where $\mathbf{\bar x}$ represents either $\hat{\mathbf x}_t^{f}$ or $\hat{\mathbf x}_t^{b}$, $\mathbf{P}_x$ is the corresponding covariance matrix for filtering or smoothing, $\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_i$ is the $i^{th}$ column of the scaled matrix square root of $\mathbf P_x$, and $n_x$ is the dimension of the state space. The total number of sigma points is $2n_x+1$. The scaling parameter, $\lambda$, is defined as 
\begin{equation}\label{eq:sigmapoints3}
	\lambda=\alpha^2( n_x+\kappa) - n_x, 
\end{equation}
\parham{(reviewer 2, comment 7:) The positive constant $\alpha$, that determines the spread of the sigma points around $\mathbf{\bar x}$, is typically set arbitrary small (e.g., $10^{-3}$) to minimise higher order effect \citep{Haykin2001}, and  $\kappa$ is an additional scaling parameter which was set to $3-n_x$ in accordance to \citet{Julier2002a}.}

The sigma vectors are propagated through the system equations and weighted to form the predicted mean and covariance. The weights are calculated by 
\begin{align}
	\mathbf W_0^{(m)}&=\frac{\lambda}{ n_x+\lambda} \\
	\mathbf W_0^{(c)}&=\frac{\lambda}{ n_x+\lambda}+(1-\alpha^2+\beta) \\
	\mathbf W_i^{(m)}&=\mathbf W_i^{(c)}=\frac{1}{2( n_x+\lambda)} \quad i=1, \dots, 2n_x, 
\end{align}
where the superscripts $m$ and $c$ denote mean and covariance and $\beta$ incorporates prior knowledge of the distribution of the states, $\mathbf{x}$ \dean{(for a Gaussian disturbance $\beta$ should be set to 2)}. Since the observation equation is linear (equation~\ref{ObservationEquation}), the standard Kalman Filter update equations are used to correct the predicted states. The state estimates from forward filtering are used to form a new set of sigma points for the smoother, as described above. The cross-covariance matrix of the states, $\mathbf M$, is also required for computing the smoother gain, as described in the summary of the steps in the URTSS algorithm is given in Table~\ref{tab:UKFAlgorithm}.

@@

\dean{(Reviewer 2, comment 6:) The work of \citet{Freeman2000} provides direct evidence that the electrode spatial frequency of the neural field is in the range of 0.1-0.4 cycles/mm and spacing required to prevent aliasing is approximately 1.25 mm. Micro-electrodes systems that satisfy this requirement are now currently in for brain computer interface and epilepsy monitoring applications~\citep{Cheung2007}, which suggests that the framework can be applied to humans with existing technology.}
